---
title: Chapter 3 Problems
output: html_document
---

```{r globe-tossing}
p_grid <- seq(from = 0, to = 1, length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, size = 9, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)

set.seed(100)
samples <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)
```

**3E1.**
```{r}
sum(samples < 0.2) / length(samples)
```
**3E2.**
```{r}
sum(samples > 0.8) / length(samples)
```
**3E3.**
```{r}
sum(samples >= 0.2 & samples <= 0.8) / length(samples)
```
**3E4.**
```{r}
quantile(samples, 0.2)
```
**3E5.**
```{r}
quantile(samples, 0.8)
```
**3E6.**
```{r, message=FALSE, warning=FALSE}
library(rethinking)
HPDI(samples, prob = 0.66)
```
**3E7.**
```{r}
PI(samples, prob = 0.66)
```
**3M1.**
```{r}
likelihood <- dbinom(8, size = 15, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```
**3M2.**
```{r}
samples <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)
HPDI(samples, prob = 0.9)

dens(samples)
```
**3M3.**
```{r}
posterior_predictive <- rbinom(1e4, size = 15, prob = samples)
(prob_8_in_15 <- sum(posterior_predictive == 8) / length(posterior_predictive))
```
**3M4.**
```{r}
post_pred <- rbinom(1e4, size = 9, prob = samples)
(prob_6_in_9 <- sum(post_pred == 6) / length(post_pred))
```
**3M5.**
```{r}
# Set a new prior
prior <- c(rep(0, length(p_grid) / 2), rep(2, length(p_grid) / 2))
# Re-calculate the new posterior
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)

# Sample from the posterior
samples <- sample(p_grid, size = 1e4, prob = posterior, replace = TRUE)
HPDI(samples, prob = 0.9)

# Posterior probability of 8 waters in 15 tosses
post_pred_15 <- rbinom(1e4, size = 15, prob = samples)
(prob_8_in_15 <- sum(post_pred_15 == 8) / length(post_pred_15))

# Posterior probability of 6 waters in 9 tosses
post_pred_15 <- rbinom(1e4, size = 9, prob = samples)
(prob_8_in_15 <- sum(post_pred_15 == 6) / length(post_pred_15))

# Plot density
dens(samples)
```
The better prior causes better estimates of the parameter $p$ with the same amount of data (9 tosses of the globe). For example, the 90% HPDI is much narrower with the better prior: 0.21 wide vs. 0.39 wide with the uniform prior. Furthermore, the density is also narrower as can be seen in the above graphs.

**3M6.**

To answer this question, we will calculate widths of the 99% PI for several numbers of tosses ($N$s). To do this, we need to know how many water tosses there were ($W$s). Once we know $W$, we can easily calculate the 99% PI for the data ($N$, $W$) with the given prior. So, for a given $N$, we just loop through each possible $W$ and calculate the width of the 99% PI. Then we loop through each $N$. 

```{r cache=TRUE}
# For a given set of 
N <- seq(from = 1, to = 5000, by = 100)
p_grid <- seq(from = 0, to = 1, length.out = 1e4)
prior <- rep(1, 1e4)
pi_width_by_toss <- function(N, p_grid, prior) {
    # Possible Ws
    W <- seq(from = 0, to = N)
    # For a given W, calculate the width 99% PI
    pi_width <- function(W) {
        l <- dbinom(W, size = N, prob = p_grid)
        post <- l * prior
        post <- post / sum(post)
        samples <- base::sample(p_grid, size = 1e4, replace = TRUE, prob = post)
        return(diff(PI(samples, prob = 0.99)))
    }
    # Calculate the widths of the 99% PI for each possible W
    return(sapply(W, pi_width, USE.NAMES = FALSE))
}

all_widths <- lapply(N, pi_width_by_toss, p_grid, prior)
all_Ns <- unlist(sapply(N, \(x){rep(x, x + 1)}))
all_widths_v <- unlist(all_widths)
plot(all_Ns, all_widths_v)
abline(h = 0.05, col = "red", lty = 2)

# Calculate the percent of widths < 0.05 per number of tosses
pct_small_widths <- sapply(all_widths, \(x) {mean(x <= 0.05)})
plot(N, pct_small_widths, type = "o")
```
Based on the chart above. It takes about 3000 tosses before we are guaranteed to have 99% PI whose width is less than 0.05; it takes about 2000 tosses for a 50% chance (depending on the observed data).
 
**3H1.**
Let $B$ be the number of boys born in $N$ births and $p$ the probability of a birth being a boy. Then, the model we will use is 
$$
\begin{align*}
B &\sim \text{Binomial}(N, p) \\
p &\sim \text{Uniform}(0,1)
\end{align*}
$$

```{r}
# Load the data
data(homeworkch3)
n_births <- length(birth1) + length(birth2)
n_boys <- sum(birth1) + sum(birth2)
# Use the same `p_grid` from earlier
prior <- rep(1, length(p_grid))
likelihood <- dbinom(n_boys, size = n_births, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
p_grid[which.max(posterior)]
```

**3H2.**
```{r}
samples <- sample(p_grid, 1e4, replace = TRUE, prob = posterior)
HPDI(samples, prob = 0.5)
HPDI(samples, prob = 0.89)
HPDI(samples, prob = 0.97)
```

**3H3.**
```{r}
post_pred <- rbinom(1e4, size = 200, prob = samples)
# Distribution of posterior predictions
dens(post_pred)
# Number of boys in the sample and the posterior predictive mean plotted as a
# vertical line
abline(v = c(n_boys, mean(post_pred)), col = c(rangi2, "red"), lty = 2)
```
The actual number of boys in the sample is close to the posterior predictive distribution, but the mean of the distribution is a bit larger.

**3H4.**
```{r}
post_pred_100 <- rbinom(1e4, size = 100, prob = samples)
dens(post_pred_100)
abline(v = sum(birth1), col = rangi2, lty = 2)
```
This model appears to overestimate the number of firstborn boys. 

**3H5.**
```{r}
births_after_female_1st_borns <- birth2[birth1 == 0]
post_pred_after_female <- rbinom(1e4,
                                 size = length(births_after_female_1st_borns),
                                 prob = samples)
dens(post_pred_after_female)
abline(v = sum(births_after_female_1st_borns), col = rangi2, lty = 2)
```





