---
title: Chapter 4 Problems and Notes
output: html_document
---

## Notes

### Example normal model of height (sec. 4.3)
```{r}
library(tidyverse)
library(rethinking)

# Load data from rethinking package and limit to adults
data(Howell1)
d2 <- Howell1[Howell1$age >= 18, ]

# Height model
mu_list <- seq(from = 150, to = 160, length.out = 100)
sigma_list <- seq(from = 7, to = 9, length.out = 100)
post <- expand.grid(mu = mu_list, sigma = sigma_list)
post$LL <- sapply(1:nrow(post), 
                  function(i) 
                    sum(dnorm(d2$height, post$mu[[i]], post$sigma[[i]], log = TRUE)))
post$prod <- post$LL +
  dnorm(post$mu, 178, 20, log = TRUE) +
  dunif(post$sigma, 0, 50, log = TRUE)
post$prob <- exp(post$prod - max(post$prod))

# Sample from posterior
sample_rows <- sample(1:nrow(post) - 1, size = 1e4, replace = TRUE, prob = post$prob)
sample_mu_sigma <- post[sample_rows, c("mu", "sigma")]
plot(sample_mu_sigma, cex = 0.5, pch = 16, col = col.alpha(rangi2, 0.1))

```

### Example linear regression (p. 97-110)
```{r}
# Scatter plot of height and weight
plot(d2$height, d2$weight)

# Define mean of weight for centering
xbar <- mean(d2$weight)

m4.3 <- quap(
    alist(
        height ~ dnorm(mu, sigma),
        mu <- a + b * (weight - xbar),
        a ~ dnorm(178, 20),
        b ~ dlnorm(0, 1),
        sigma ~ dunif(0, 50)
    ),
    data = d2
)

# Summary of posterior
precis(m4.3)

# Sample from the posterior
post <- extract.samples(m4.3)
# Mean of posterior a and b parameters
a_map <- mean(post$a)
b_map <- mean(post$b)
ggplot(data = d2, aes(x = weight, y = height)) +
  geom_point(color = "red", shape = "o", size = 4) +
  geom_abline(slope = b_map, intercept = a_map - b_map * xbar, lwd = 2) +
  geom_abline(data = post[1:20, ],
              aes(slope = b, intercept = a - b * xbar),
              alpha = 0.4)

# Distribution of mu at weight values
weight_seq <- 25:70
mu <- link(m4.3, data = data.frame(weight = weight_seq))
mu_mean <- apply(mu, 2, mean)
mu_pi <- apply(mu, 2, PI, prob = 0.89)
# Reshape `mu_pi` for easier plotting
mu_pi <- t(mu_pi) %>%
  as_tibble() %>%
  rename_with(~{paste("mu", .x, sep = "_")}) %>%
  mutate(weight = weight_seq)
# Generate posterior predictions
sim_height <- sim(m4.3, data = list(weight = weight_seq), n = 1e4)
height_pi <- apply(sim_height, 2, PI, prob = 0.89) %>%
  t() %>%
  as_tibble() %>%
  rename_with(~{paste("height", .x, sep = "_")}) %>%
  mutate(weight = weight_seq)


ggplot() +
  geom_point(data = d2,
             aes(x = weight, y = height),
             color = "red",
             shape = "o",
             size = 4) +
  geom_abline(slope = b_map, intercept = a_map - b_map * xbar) +
  geom_ribbon(data = mu_pi, aes(x = weight, ymin = `mu_5%`, ymax = `mu_94%`),
              alpha = 0.3,
              fill = rangi2) +
  geom_ribbon(data = height_pi,
              aes(x = weight, ymin = `height_5%`, ymax = `height_94%`),
              alpha = 0.3) +
  theme_minimal()
```

## Problems
**4E1.** The first line $y_i \sim \text{Normal}(\mu,\sigma)$ is the likelihood.

**4E2.** There are two paramaters in the posterior distribution $\mu$ and $\sigma$.

**4E3.** Bayes' Theorem says
$$
\begin{align*}
\text{Pr}(\mu,\sigma|y_i)&\propto\text{Pr}(y_i|\mu,\sigma)\text{Pr}(\mu)\text{Pr}(\sigma) \\
&=\frac{\text{Normal}(y_i|\mu,\sigma)\text{Normal}(\mu|0,10)\text{Exponential}(\sigma|1)}{\int\text{Normal}(y_i|\mu,\sigma)\text{Normal}(\mu|0,10)\text{Exponential}(\sigma|1)\text d\mu\text d\sigma}
\end{align*}
$$

**4E4.** The linear model is the second line $\mu_i=\alpha + \beta x_i$.

**4E5.** There are three parameters in the posterior distribution: $\alpha$, $\beta$, and $\sigma$.

**4M1.** 
```{r}
prior_mu <- rnorm(1e4, mean = 0, sd = 10)
prior_s <- rexp(1e4, rate = 1)
y_prior_pred <- rnorm(1e4, mean = prior_mu, s = prior_s)
hist(y_prior_pred)
```

**4M2.**
```{r eval=FALSE}
quap(
    alist(
        y ~ dnorm(mu, sigma),
        mu ~ dnorm(0, 10),
        sigma ~ dexp(1)
    )
)
```

**4M3.**
$$
\begin{align*}
y_i &\sim \text{Normal}(\mu, \sigma) \\
\mu_i &= \alpha + \beta x_i \\
\alpha &\sim \text{Normal}(0, 10) \\
\beta &\sim \text{Uniform}(0, 1) \\
\sigma &\sim \text{Exponential}(1)
\end{align*}
$$

**4M4.**
