---
title: Chapter 13 Problems and Notes
output: html_document
---

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(rethinking)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Practice

**13E1.**

The $\alpha_{TANK}\sim\text{Normal}(0,1)$ will produce more shrinkage than the prior with a standard deviation of 2. The smaller standard deviation is more skeptical of outlying values, so will be less likely to overfit them.

**13E2.**

$$\begin{align}
y_i &\sim \text{Binomial}(1, p_i) \\
\text{logit}(p_i) &= \alpha_{GROUP[i]} + \beta x_i\\
\alpha_{GROUP} &\sim \text{Normal}(\bar\alpha, \sigma) \\
\beta &\sim \text{Normal}(0, 0.5) \\
\bar\alpha &\sim \text{Normal}(0, 0.5) \\
\sigma &\sim \text{Exponential}(1)
\end{align}$$

**13E3.**

$$\begin{align}
y_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha_{GROUP[i]} + \beta x_i\\
\alpha_{GROUP} &\sim \text{Normal}(\bar\alpha, \tau) \\
\beta &\sim \text{Normal}(0, 0.5) \\
\bar\alpha &\sim \text{Normal}(0, 0.5) \\
\tau &\sim \text{Exponential}(1) \\
\sigma &\sim \text{Exponential}(1)
\end{align}$$

**13E4.**

$$\begin{align}
y_i &\sim \text{Poisson}(\lambda) \\
\log(\lambda) &= \alpha_j \\
\alpha_j &\sim \text{Normal}(\bar\alpha, \sigma) \\
\bar\alpha &\sim \text{Normal}(0, 1) \\
\sigma &\sim \text{Exponential}(1)
\end{align}$$

**13E5.**

$$\begin{align}
y_i &\sim \text{Poisson}(\lambda) \\
\log(\lambda) &= \alpha_j + \beta_j \\
\alpha_j &\sim \text{Normal}(\bar\alpha, \sigma) \\
\beta_j &\sim \text{Normal}(0, \tau) \\
\bar\alpha &\sim \text{Normal}(0, 1) \\
\sigma &\sim \text{Exponential}(1) \\
\tau &\sim \text{Exponential}(1)
\end{align}$$

**13M1.**

Load the data. Fit the model from the chapter to confirm everything is working.
```{r}
data(reedfrogs)
d <- reedfrogs |>
  mutate(tank = row_number(),
         pr = as.integer(pred),
         sz = as.integer(size))
dat <- list(
  s = d$surv,
  n = d$density,
  tank = d$tank,
  pr = d$pred,
  sz = d$size
)
m13_2 <- ulam(
  alist(
    s ~ dbinom(n, p),
    logit(p) <- a[tank],
    a[tank] ~ dnorm(a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat,
  chains = 4,
  cores = 4,
  log_lik = TRUE
)
precis(m13_2)
```

Now add `predation` and `size` in various configurations.
```{r}
# Model adding predation only
m_p <- ulam(
  alist(
    s ~ dbinom(n, p),
    logit(p) <- a[tank] + bp[pr],
    a[tank] ~ dnorm(a_bar, sigma),
    bp[pr] ~ dnorm(0, 1),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat,
  chains = 4,
  cores =4,
  log_lik = TRUE
)
precis(m_p, depth = 2, pars = c("bp", "a_bar", "sigma"))
# Model adding size only
m_s <- ulam(
  alist(
    s ~ dbinom(n, p),
    logit(p) <- a[tank] + bs[sz],
    a[tank] ~ dnorm(a_bar, sigma),
    bs[pr] ~ dnorm(0, 1),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat,
  chains = 4,
  cores =4,
  log_lik = TRUE
)
precis(m_s, depth = 2, pars = c("bs", "a_bar", "sigma"))
# Model with interaction of predation and size
m_sp <- ulam(
  alist(
    s ~ dbinom(n, p),
    logit(p) <- a[tank] + b[pr, sz],
    a[tank] ~ dnorm(a_bar, sigma),
    matrix[pr, sz]:b ~ normal(0, 1),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat,
  chains = 4,
  cores =4,
  log_lik = TRUE
)
precis(m_sp, depth = 3, pars = c("b", "a_bar", "sigma"))
```

The mean values of $\sigma$ vary across the four models. It is the smallest in the model with the interaction of `size` and `pred`, slightly larger in the model with only `pred`, and then almost double the size in the model without any effects or just `size`. The reason is that predation has a large effect on survival. In models that don't include it, the population-level variance has to include it.

```{r}
coeftab_plot(coeftab(m13_2, m_p, m_s, m_sp), pars = "sigma")
```

**13M2.**

```{r}
compare(m13_2, m_p, m_s, m_sp)
```

**13M3.**

```{r}
m13m3 <- ulam(
  alist(
    s ~ dbinom(n, p),
    logit(p) <- a[tank],
    a[tank] ~ dcauchy(a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat,
  chains = 4,
  cores = 4,
  log_lik = TRUE
)

precis(m13m3, depth = 2)
```

Now we compare to model `m13_2` which was fit in problem 13M1.
```{r}
# Can't get coeftab to work with the a parameters, so need to compare manually
coef_table <- tibble(
  "par" = precis(m13_2, depth = 2)@row.names,
  "m13_2" = precis(m13_2, depth = 2)@.Data[[1]]
) |>
left_join(
  tibble(
  "par" = precis(m13m3, depth = 2)@row.names,
  "m13m3" = precis(m13m3, depth = 2)@.Data[[1]]
  ),
  by = "par"
) |>
mutate(tank = as.integer(str_extract(par, "[0-9]{1,2}")),
       # Convert a parameters to probability scale
       m13_2 = if_else(str_detect(par, "^a"), inv_logit(m13_2), m13_2),
       m13m3 = if_else(str_detect(par, "^a"), inv_logit(m13m3), m13m3)) |>
left_join(d[, c("tank", "propsurv")], by = "tank") |>
rename(actual = propsurv) |>
# Calculate overall weighted average survival of the sample
mutate(actual = if_else(par == "a_bar", sum(d$surv) / sum(d$density), actual))

overall_means <- coef_table |>
  select(-tank) |>
  filter(par == "a_bar") |>
  pivot_longer(-par, names_to = "model", values_to = "post_mean")

coef_table |>
  filter(str_detect(par, "a\\[")) |>
  pivot_longer(-c("par", "tank"),
               names_to = "model",
               values_to = "post_mean") |>
ggplot(aes(x = reorder(tank, -tank), y = post_mean, color = model)) +
geom_point() +
geom_hline(data = overall_means, 
           aes(yintercept = post_mean, color = model), linetype = 2) +
coord_flip() +
labs(x = "Tank", y = "Proportion Survival", color = "") +
theme_classic()
```

**13M4.**

```{r}
m13m4 <- ulam(
  alist(
    s ~ dbinom(n, p),
    logit(p) <- a[tank],
    a[tank] ~ dstudent(2, a_bar, sigma),
    a_bar ~ dnorm(0, 1.5),
    sigma ~ dexp(1)
  ),
  data = dat,
  chains = 4,
  cores = 4,
  log_lik = TRUE
)

# Add this model to the coef_table from the last problem.
coef_table <- coef_table |>
  left_join(tibble(
    "par" = precis(m13m4, depth = 2)@row.names,
    "m13m4" = precis(m13m4, depth = 2)@.Data[[1]]
  ), by = "par") |>
  mutate(m13m4 = if_else(str_detect(par, "^a"), inv_logit(m13m4), m13m4))

overall_means <- coef_table |>
  select(-tank) |>
  filter(par == "a_bar") |>
  pivot_longer(-par, names_to = "model", values_to = "post_mean")

coef_table |>
  filter(str_detect(par, "a\\[")) |>
  pivot_longer(-c("par", "tank"),
               names_to = "model",
               values_to = "post_mean") |>
ggplot(aes(x = reorder(tank, -tank), y = post_mean, color = model)) +
geom_point() +
geom_hline(data = overall_means, 
           aes(yintercept = post_mean, color = model), linetype = 2) +
coord_flip() +
labs(x = "Tank", y = "Proportion Survival", color = "") +
theme_classic()
```
