---
title: Solutions for Statistical Rethinking 2023 Homework Week 03
output: html_document
--- 
[Link to problem sheet](https://github.com/rmcelreath/stat_rethinking_2023/blob/main/homework/week03.pdf) 

```{r setup, echo=FALSE, message=FALSE}
library(rethinking)
library(dagitty)
library(tidyverse)

data(foxes)
```

> 1. The first two problems are based on the same data. The data in `data(foxes)` are 116 foxes from 30 different urban groups in England. These fox groups are like street gangs. Group size (`groupsize`) varies from 2 to 8 individuals. Each group maintains its own (almost exclusive) urban territory. Some territories are larger than others. The `area` variable encodes this information. Some territories also have more `avgfood` than others. And food influences the `weight` of each fox. Assume this DAG:
```{r echo=FALSE}
fox_dag <- dagitty('dag{
    A [pos="0, -0.5"]
    F [pos="-0.5, 0"]
    G [pos="0.5, 0"]
    W [pos="0, 0.5"]
    A -> F
    F -> G
    F -> W
    G -> W
    }')
drawdag(fox_dag)
```

> where $F$ is `avgfood`, $G$ is `groupsize`, $A$ is `area`, and $W$ is `weight`.
> 
> Use the backdoor criterion and estimate the total causal influence of $A$ on $F$. What effect would increasing the area of a territory have on the amount of food inside it?

The influence of $A$ on $F$ has only one path, the direct one. Therefore, there are no backdoors to consider, and we can get the total influence with just regressing $F$ on $A$. 

```{r}
# Create dataset of variables and standardize
d <- data.frame(a = standardize(foxes$area),
                f = standardize(foxes$avgfood),
                g = standardize(foxes$groupsize),
                w = standardize(foxes$weight))

# Fit a model
m1 <- quap(
    alist(
        f ~ dnorm(mu, sigma),
        mu <- c + b * a,
        c ~ dnorm(0, 0.2), # tight around zero for standardized variables
        b ~ dnorm(0, 1), # should be positive, but not sure
        sigma ~ dexp(1)
    ),
    data = d
)

precis(m1)

post <- extract.samples(m1)
dens(post$b)
```

According to this model and these data, increasing the area of the group by one standard deviation would increase the amount of food between 0.8 and 0.95 standard deviations. 

> 2. Infer the **total** causal effect of adding food $F$ to a territory on the weight $W$ of foxes. Can you calculate the causal effect by simulating an intervention on food?

According to the DAG, $F$ affects $W$ through two paths: directly and through $G$. There are no backdoors from $F$ to $W$, so there is nothing else to condition on. 

```{r}
m2 <- quap(
    alist(
        w ~ dnorm(mu, sigma),
        mu <- c + bf * f,
        c ~ dnorm(0, 0.2),
        bf ~ dnorm(0, 0.5),
        sigma ~ dexp(1)
    ),
    data = d
)

precis(m2)

post <- extract.samples(m2)
dens(post$bf)
```

There does not appear to be much of an effect of $F$ on $W$ based on this model and these data. 

> 3. Infer the **direct** causal effect of adding food $F$ to a territory on the weight $W$ of foxes. In light of your estimates from this problem and the previous one, what do you think is going on with these foxes?

We can calculate the direct effect of adding food by conditioning on $G$:

```{r}
m3 <- quap(
    alist(
        ## F -> W <- G
        w ~ dnorm(mu, sigma),
        mu <- c + bf * f + bg * g,
        c ~ dnorm(0, 0.2),
        bf ~ dnorm(0, 0.5),
        bg ~ dnorm(0, 0.5),
        sigma ~ dexp(1),
    ),
    data = d
)

precis(m3)
```

Based on this model, the direct effect of an increase of food is an increase in weight, but the effect of group size increase cancels it out.

I think that as more food is available in a given area, it attracts more foxes to the group. The amount of food per fox stays relatively constant so that the weight of each fox doesn't change. We can confirm this by modeling the direct effect of $F$ on $G$.

```{r}
m3_a <- quap(
    alist(
        g ~ dnorm(mu, sigma),
        mu <- cg + bg * f,
        cg ~ dnorm(0, 0.2),
        bg ~ dnorm(0, 0.5),
        sigma ~ dexp(1)
    ),
    data = d
)
precis(m3_a)
```

This model clearly shows that an increase in $F$ is strongly associated with an increase in $G$, confirming our story.

> 4. Suppose there is an unobserved confound that influences $F$ and $G$, like this:

```{r echo=FALSE}
fox_dag2 <- dagitty('dag{
    A [pos="-0.5, -0.5"]
    F [pos="-0.5, 0"]
    G [pos="0.5, 0"]
    W [pos="0, 0.5"]
    U [unobserved, pos="0.5, -0.5"]
    A -> F
    F -> G
    U -> F
    U -> G
    F -> W
    G -> W
    }')
drawdag(fox_dag2)
```

> Assuming the DAG above is correct, again estimate the total and direct causal effects of $F$ on $W$. What impact does the unobserved confound have?

Apparently, this is a trick question. We can't find the total effect because we would have to condition on $U$, which is unobservable. The direct effect of $F$ on $W$ has already been estimated in #3.