---
title: Solutions for Statistical Rethinking 2023 Homework Week 01
output: html_document
--- 
[Link to problem sheet](https://github.com/rmcelreath/stat_rethinking_2023/blob/main/homework/week01.pdf) 

> 1. Suppose the globe tossing data (Lecture 2, Chapter 2) had turned out to be 4 water and 11 land. Construct the posterior distribution.
```{r p1}
# Define the possible values of $p$
prior <- seq(from = 0, to = 1, by = 0.25)
# Define the data
data <- c("W" = 4, "L" = 11)
# Calculate the posterior and then normalize
post <- sapply(prior, \(p) p^data["W"] * (1 - p)^data["L"])
(post <- post / sum(post))
# Plot the posterior
barplot(post, names.arg = prior)
```

> 2. Using the posterior distribution from **1**, compute the posterior predictive distribution for the next 5 tosses of the same globe. I recommend you use the sampling method.

```{r p2}
# Simulate $p$ values from the posterior
sim_p <- sample(prior, size = 1e4, replace = TRUE, prob = post)
# Simulate 5 tosses of the globe using probabilities from `sim_p`
sim_toss <- sapply(sim_p,
                   \(p) sample(c("W", "L"), size = 5, replace = TRUE,
                               prob = c(p, 1 - p))
                  )
# Plot the distribution for the number of Ws from the simulated tosses
n_w <- apply(sim_toss, 2, \(toss) sum(toss == "W"))
barplot(table(n_w) / length(n_w))
```

> 3. Use the posterior predictive distribution from **2** to calculate the probability of 3 or more water samples in the next 5 tosses.

```{r p3}
(sum(n_w >= 3) / length(n_w))
```

> 4. Suppose you observe W = 5 water points, but you forgot to write down how many times the globe was tossed, so you don't know the number of land points $L$. Assume $p = 0.7$ and compute the posterior distribution of the number of tosses $N$.

*This solution seems too complicated, but I think it works.*

We are after

$$\mathbb P(N=n | W=5, p=0.7)$$

By Bayes' Theorem, this is equal to

$$\mathbb P(N=n | W=5, p=0.7) = \frac{\mathbb P(W = 5 | N=n, p=0.7)\mathbb P(N=n, p=0.7)}{\mathbb P(W = 5 | p = 0.7)}$$.

The difficulty is "prior" on $N$: $\mathbb P(N=n|p=0.7)$. But, let's suppose we also know that $N<=M-1$ for some integer $m$. Then we can put a uniform prior on $N$, so that $\mathbb P(N=n|p=0.7)=1/m$ (assuming $N\in\{0,1,\dots,m-1\}$). Then, we get

$$\begin{align}
\mathbb P(N=n | W=5, p=0.7) &= \frac{\mathbb P(W = 5 | N=n, p=0.7)\mathbb P(N=n, p=0.7)}{\mathbb P(W = 5 | p = 0.7)}\\
&=\frac{\mathbb P(W = 5 | N=n, p=0.7)\times \frac1m}{\sum_{k=0}^{m-1}\mathbb P(W = 5 | N=k, p=0.7)\times\frac 1m}\\
&=\frac{\mathbb P(W = 5 | N=n, p=0.7)}{\sum_{k=0}^{m-1}\mathbb P(W = 5 | N=k, p=0.7)}
\end{align}
$$

Now we can take the limit as $m\rightarrow\infty$. We can approximate the denominator, and the numerator can be calculated with `dbinom()`.

```{r p4}
p <- 0.7
w <- 5
# Values of N
ns <- seq(from = 0, to = 1000)
# Approximate the denominator
approx_denom <- sum(dbinom(w, ns, p))
# Posterior distribution for N from 0 to 1000
post_n <- dbinom(w, ns, p) / approx_denom
# Plot posterior
barplot(post_n, names.arg = ns)
# Zoomed in plot for posterior of N from 1 to 20
barplot(post_n[1:20], names.arg = ns[1:20])
```